---
title: "Research Analysis Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
#| label: load-data

load(here("data", "data_constructed", "analytical_data.RData"))
```

# Intermarriage benchmarks

```{r}
#| label: fig-intermar-exogamy-compare
#| fig-cap: Relationship between the exogamy parameter for the simulation and resulting log odds ratio of intergroup marriage, based on benchmark simulations where each endogamy parameter was allowed to operate for a 15 year period. The even benchmark was for a population evenly split between groups while the uneven benchmark applied to a population that was split 80/20 by group. The grey line fits a line with a slope of two and an intercept of zero. 

ggplot(intermar, aes(x = exogamy, y = lor, 
                     group = group_share, color = group_share))+
  geom_abline(intercept = 0, slope =2, size = 1.2, color = "grey")+
  geom_point(alpha = 0.5)+
  geom_smooth(se = FALSE, alpha = 0.5)+
  labs(x = "exogamy parameter", y = "log odds ratio of intermarriage",
       color = "group share")+
  theme_bw()
```

@fig-intermar-exogamy-compare plots the exogamy parameter of each simulation period in the benchmark tests versus the actual log odds ratio of group intermarriage in that same period. As I expected, the log odds ratio used in the simulation is almost exactly half the odds of the cross product log odds ratio. That linear relationship may start to moderate at the most extreme values, but in general its a very good fit. It also produces the exact same estimates regardless of group size. 