---
title: "Organize ACS/Census Data Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

This script will read in Census and ACS data from 1960 to 2019 to construct broad interracial marriage patterns over that entire time period. We will use the imputation of Hispanicity in the 1960 data.

```{r}
#| label: read-data

census <- read_fwf(here("data", "data_raw", "acs_census", "usa_00138.dat.gz"),
                   col_positions = fwf_cols(YEAR = c(1,4),
                                            SAMPLE = c(5, 10),
                                            HHWT = c(32, 41),
                                            SEX = 82,
                                            AGE = c(83, 85),
                                            RACED = c(88, 90),
                                            HISPAN = 91,
                                            HISP1970 = 103,
                                            BPL = c(95, 97),
                                            SEX_SP = 104,
                                            AGE_SP = c(105, 107),
                                            RACED_SP = c(109, 111),
                                            HISPAN_SP = 112,
                                            HISP1970_SP = 119,
                                            BPL_SP = c(116, 118)),
                   col_types = cols(.default = "i", HHWT = "d"))
```

In more recent data we will have some same sex couples. Since we are interested in genealogies, we want opposite sex couples. We also want to have a consistent comparison over time. So, we will exclude same sex couples.

```{r}
#| label: check-spouse-sex

# 
table(census$SEX, census$SEX_SP)

census <- census |>
  filter(SEX_SP == 2)

table(census$SEX, census$SEX_SP)
```

Replace HISPAN with HISP1970 for the 1970 data.

```{r}
#| label: restrict-1970

census <- census |>
  mutate(HISPAN = ifelse(YEAR == 1970, HISP1970, HISPAN),
         HISPAN_SP = ifelse(YEAR == 1970, HISP1970_SP, HISPAN_SP))
```

Husbands should be restricted to native born. Lets also restrict wives.

```{r}
#| label: restrict-nativity

table(census$BPL)

census <- census |>
  filter(BPL_SP <= 99)

table(census$BPL)
```

Husbands should be restricted to 25-34. Lets check the ages of men and women.

```{r}
#| label: check-age

summary(census$AGE)
summary(census$AGE_SP)
```

Hmmm, a bit troubling on the minimum values. Data errors? I will leave them in for now.

Ok, now we are ready to do the data coding/cleaning.

```{r}
#| label: create-vars

# lets create a function for assigning race, since we will do it for both spouses
code_race <- function(raced, hispan) {
  
  race <- case_when(
    hispan != 0 | raced == 110 ~ "Hispanic",
    raced == 100 ~ "White",
    raced == 200 ~ "Black",
    raced >= 300 & raced <= 399 ~ "AIAN",
    # we have to do API because 1980 and 1990
    (raced >= 400 & raced <= 699) | (raced >= 860 & raced <= 869) ~ "API",
    raced == 700 ~ "Other", 
    # lets do major biracial groups for fun
    raced == 801 ~ "White/Black",
    raced == 802 ~ "White/AIAN",
    raced >= 803 & raced <= 825 ~ "White/API",
    raced == 830 ~ "Black/AIAN",
    raced >= 831 & raced <= 842 ~ "Black/API",
    TRUE ~ "Other Multiracial")
  
  factor(race, 
         levels = c("White", "Black", "AIAN", "Hispanic", "API", "Other",
                    "White/Black", "White/AIAN", "White/API", 
                    "Black/AIAN", "Black/API", "Other Multiracial"))
}

census <- census |>
  mutate(race_husband = code_race(RACED, HISPAN),
         race_wife = code_race(RACED_SP, HISPAN_SP), 
         year = YEAR) |>
  # remove hispanics in 1960 and 1970 due to measurement issues
  filter(!(race_husband == "Hispanic" & year <= 1970))
  

census <- census

table(census$race_husband)
table(census$race_wife)
```


```{r}
#| label: create-master-tab
tab_mar <- table(census$race_husband, census$race_wife, census$year)
```

```{r}
calculate_or <- function(group1, group2) {
  lor <- log(tab_mar[group1, group2,])+log(tab_mar[group2, group1,])-
    log(tab_mar[group1, group1,])-log(tab_mar[group2, group2,])
  se <- sqrt(1/tab_mar[group1, group2,]+1/tab_mar[group2, group1,]+
               1/tab_mar[group1, group1,]+1/tab_mar[group2, group2,])
  tibble(year = as.numeric(names(lor)), lor, se, 
         type = paste(group1, group2, sep="-"))
}
```


```{r}
#| label: get-lor-df

all_combos <- tribble(
  ~race1, ~race2,
  "White", "Black",
  "White", "AIAN",
  "White", "API",
  "White", "Hispanic",
  "Black", "AIAN",
  "Black", "API",
  "Black", "Hispanic",
  "AIAN", "API",
  "AIAN", "Hispanic",
  "API", "Hispanic",
  "White/Black", "White",
  "White/Black", "Black",
  "White/API", "White",
  "White/API", "API",
  "White/AIAN", "White",
  "White/AIAN", "AIAN"
)

lor <- pmap(all_combos, function(race1, race2) {
  calculate_or(race1, race2)
  }) |>
  bind_rows() |>
  filter(lor > -Inf)
```

```{r}
#| label: fig-trends
#| fig-cap: Trends in interracial marriage involving a White or part-White partner
#| layout-ncol: 2
#| fig-subcap:
#|   - Single race groups
#|   - White/Black
#|   - White/API
#|   - White/AIAN

lor |>
  filter(type %in% c("White-Black","White-AIAN","White-API", "White-Hispanic")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()

lor |>
  filter(type %in% c("White-Black","White/Black-White","White/Black-Black")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()

lor |>
  filter(type %in% c("White-API","White/API-White","White/API-API")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()

lor |>
  filter(type %in% c("White-AIAN","White/AIAN-White","White/AIAN-AIAN")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()
```

```{r}
#| label: fig-trends-nonwhite
#| fig-cap: Trends in interracial marriage by focal minority group
#| layout-ncol: 2
#| fig-subcap:
#|   - Black partner
#|   - AIAN partner
#|   - API partner
#|   - Hispanic partner

lor |>
  filter(type %in% c("White-Black", "Black-AIAN","Black-API","Black-Hispanic")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()

lor |>
  filter(type %in% c("White-AIAN", "Black-AIAN", "AIAN-API", "AIAN-Hispanic")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()

lor |>
  filter(type %in% c("White-API", "Black-API","AIAN-API","API-Hispanic")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()

lor |>
  filter(type %in% c("White-Hispanic", "Black-Hispanic","AIAN-Hispanic","API-Hispanic")) |>
  ggplot(aes(x = year, y = lor, group = type))+
  geom_ribbon(aes(ymin = lor - 1.96 * se, ymax = lor + 1.96 * se, fill = type),
              alpha = 0.3)+
  geom_smooth(se = FALSE, aes(color = type))+
  geom_point(aes(color = type), alpha = 0.5)+
  theme_bw()
```