---
title: "Supplementary Materials"
format:
  pdf:
    keep-tex: false
    fig-pos: "!t"
    include-in-header: 
      text: |
        \usepackage{dcolumn}
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data", "data_constructed", "analytical_data.RData"))

years_runin <- 100
```

```{r}
#| label: set-palettes

palette_scenario <- c(
  "absorbing state"      = "#E7298A",  
  "high peak, resilient" = "#2C7FB8",  
  "high peak, decay"     = "#66C2A5", 
  "low peak, resilient"  = "#D95F02", 
  "low peak, decay"      = "#FDB462"  
)

palette_group <- c(
  "A"  = "#000080",
  "B"  = "#FF6347",
  "AB" = "#FFD700"
)

palette_intermar <- c(
  "B-AB"  = "#00A6A6",
  "A-AB"  = "#F18F01",
  "A-B"   = "#5F0F40"
)

palette_mar_growth <- c(
  "progress" = "#1B4332",
  "plateau"  = "#C9184A"
)
```


```{r}
#| label: clean-up-labels

sims_pop <- sims_pop |>
  mutate(
    group = factor(group, levels = 1:3, labels = c("A", "B", "AB")),
    pop_share = factor(pop_share, levels = c("uneven", "even"), 
                       labels = c("uneven (85/15)", "even (50/50)")),
    scenario = str_replace(scenario, "_", " peak, "),
    scenario = str_replace(scenario, "total peak, rigid", "absorbing state"),
    scenario = factor(
      scenario,
      levels = c("absorbing state", "high peak, resilient", "high peak, decay",
                 "low peak, resilient", "low peak, decay")))

sims_mar <- sims_mar |>
  mutate(
    hgroup = factor(hgroup, levels = 1:3, labels = c("A", "B", "AB")),
    wgroup = factor(wgroup, levels = 1:3, labels = c("A", "B", "AB")),
    pop_share = factor(pop_share, levels = c("uneven", "even"), 
                       labels = c("uneven (85/15)", "even (50/50)")),
    scenario = str_replace(scenario, "_", " peak, "),
    scenario = str_replace(scenario, "total peak, rigid", "absorbing state"),
    scenario = factor(
      scenario,
      levels = c("absorbing state", "high peak, resilient", "high peak, decay",
                 "low peak, resilient", "low peak, decay")))
```

```{r}
#| label: count-pop-group
#| include: false

total_pop_group <- map(0:300, function(x) {
  sims_pop |>
    get_pop_alive(x) |>
    group_by(group, pop_share, mar_growth, scenario) |>
    summarize(year = x, population = n(), .groups = "drop") |>
    group_by(pop_share, mar_growth, scenario) |>
    mutate(share = population / sum(population), .groups = "drop") |>
    arrange(pop_share, mar_growth, scenario, group)
}) |>
  bind_rows()

total_pop <- total_pop_group |>
  group_by(year, pop_share, mar_growth, scenario) |>
  summarize(population = sum(population), .groups = "drop")


total_pop_ancestry <- map(0:300, function(x) {
  sims_pop |>
    get_pop_alive(x) |>
    mutate(ancestry = case_when(
      ancestry_group1 == 1 ~ "A",
      ancestry_group2 == 1 ~ "B",
      TRUE ~ "AB"
    )) |>
    group_by(ancestry, pop_share, mar_growth, scenario) |>
    summarize(year = x, population = n(), .groups = "drop") |>
    group_by(pop_share, mar_growth, scenario) |>
    mutate(share = population / sum(population), .groups = "drop") |>
    arrange(pop_share, mar_growth, scenario, ancestry)
}) |>
  bind_rows()
```


# Testing Demographic Rates

```{r}
#| label: functions-demographic

# given a calendar year, calculate person years spent in a set of age groups

calculate_py_in_year <- function(df, year, age_group) {
  df |>
    mutate(
      # deal with missing values for living people
      dod = if_else(is.na(dod), Inf, dod),
      # age at start of year
      age_start = floor(year - dob),
      # next birthday during year
      next_bday = dob + age_start + 1,
      # times
      time_exit = pmin(dod, year + 1),
      time_enter = pmax(dob, year),
      time_total = pmax(0, time_exit - time_enter)
    ) |>
    # remove people who weren't born or died before this time
    filter(time_total > 0) |>
    # split into time spent at both ages
    mutate(
      # time before birthday
      time_lower = pmax(0, pmin(time_exit, next_bday) - time_enter),
      # time after birthday
      time_upper = pmax(0, time_exit - pmax(time_enter, next_bday)),
      age_lower = age_start,
      age_upper = age_start + 1
    ) |>
    select(age_lower, age_upper, time_lower, time_upper) |>
    # now pivot so each segment is on a separate row
    pivot_longer(cols = everything(),
                 names_to = c(".value", "part"),
                 names_pattern = "(age|time)_(lower|upper)") |>
    # I don't think this should happen, but just in case
    filter(time >= 0) |>
    # group up age and summarize time
    mutate(age_group = cut(age, breaks = age_group, right = FALSE)) |>
    # drop cases outside age range
    filter(!is.na(age_group)) |>
    group_by(age_group, .drop = FALSE) |>
    summarize(years_lived = sum(time), .groups = "drop")
}

# calculate total person years across all years where years are grouped
calculate_py <- function(df, year_group, age_group) {
  
  map(min(year_group):(max(year_group) - 1), function(x) {
    calculate_py_in_year(df, x, age_group) |>
      mutate(year = x)
  }) |>
    bind_rows() |>
    mutate(year_group = cut(year, year_group, right = FALSE)) |>
    group_by(year_group, age_group, .drop = FALSE) |>
    summarize(years_lived = sum(years_lived), .groups = "drop")
}

# count deaths by year and age group
calculate_deaths <- function(df, year_group, age_group) {

    df |>
    # remove living people
    filter(!is.na(dod)) |>
    # calculate age at death
    mutate(age_death = dod - dob) |>
    # cut age and year
    mutate(
      age_group = cut(age_death, age_group, right = FALSE),
      year_group = cut(dod, year_group, right = FALSE)
    ) |>
    select(dob, dod, age_death, age_group, year_group) |>
    group_by(year_group, age_group) |>
    summarize(deaths = n(), .groups = "drop") |>
    filter(!is.na(year_group))
}

calculate_births <- function(df, year_group, age_group) {
  
  # first attach mom information
  moms <- df |>
    select(pid, dob) |>
    rename(mom = pid, dob_mom = dob)
  
  df |> 
    # get rid of people without moms
    filter(!is.na(mom)) |>
    # add mom info
    left_join(moms, by = "mom") |>
    select(pid, sex, dob, mom, dob_mom) |>
    # calculate mom's age
    mutate(age_mom = dob - dob_mom) |>
    # cut age and year
    mutate(
      age_group = cut(age_mom, age_group, right = FALSE),
      year_group = cut(dob, year_group, right = FALSE)
    ) |>
    # count births
    group_by(year_group, age_group) |>
    summarize(births = n(), .groups = "drop") |>
    filter(!is.na(year_group))
}

# calculate life tables from a table of the format returned from functions 
# above 
get_le <- function(rates) {
  
  # get it started
  life_table <- rates |>
    rename(nmx = asmr) |>
    mutate(x = as.numeric(str_extract_all(age_group, "\\d+", simplify = TRUE)[,1]),
           n = as.numeric(str_extract_all(age_group, "\\d+", simplify = TRUE)[,2])-x,
           nax = n / 2) |>
    select(x, n, nax, nmx)
    
  # adjust for open-ended row
  life_table$n[nrow(life_table)] <- NA
  life_table$nax[nrow(life_table)] <- 1/life_table$nmx[nrow(life_table)]
  
  
  # now do the rest
  life_table <- life_table |>
    mutate(nqx = c(1 - exp(-nmx[-nrow(life_table)] * n[-nrow(life_table)]), 1),
           lx_end = cumprod(1-nqx),
           lx_start = c(1, lx_end[-nrow(life_table)]),
           ndx = lx_start-lx_end,
           nLx = ifelse(is.na(n), ndx * nax, lx_end * n + ndx * nax),
           ex = rev(cumsum(rev(nLx))))
  
  return(life_table)
}

# caculate midpoint of year_group and nicer display name
convert_year_group <- function(df) {
  
  years <- str_extract_all(df$year_group, "\\d+", simplify = TRUE)
  start <- as.numeric(years[,1])
  end <- as.numeric(years[,2])
  midpoint <- start + (end - start) / 2
  interval <- paste(start, end-1, sep = "-")
  
  
  df |>
    mutate(year_midpoint = midpoint,
           year_group = fct_reorder(interval, midpoint, max))
}
```

```{r}
#| label: calculate-demog-rates

year_group <- seq(from = 0, to = 300, by = 25)
age_group <- c(0, 1, seq(from = 5, to = 100, by = 5), 150)

person_years <- sims_pop |>
  group_by(sex, pop_share, mar_growth, scenario) |>
  group_modify(~ calculate_py(.x, year_group, age_group)) |>
  ungroup()

deaths <- sims_pop |>
  group_by(sex, pop_share, mar_growth, scenario) |>
  group_modify(~ calculate_deaths(.x, year_group, age_group)) |>
  ungroup()

death_rates <- full_join(deaths, person_years) |>
  mutate(asmr = deaths / years_lived)

life_tables <- death_rates |>
  group_by(year_group, sex, pop_share, mar_growth, scenario) |>
  group_modify(~ get_le(.x)) |>
  ungroup()

births <- sims_pop |>
  group_by(pop_share, mar_growth, scenario) |>
  group_modify(~ calculate_births(.x, year_group, age_group)) |>
  ungroup()

birth_rates <- person_years |>
  filter(sex == "Female") |>
  right_join(births) |>
  mutate(asfr = births / years_lived)

tfr <- birth_rates |>
  group_by(year_group, pop_share, mar_growth, scenario) |>
  summarize(tfr = 5 * sum(asfr), .groups = "drop")

# population growth rates at 10 year intervals
pop_growth <- total_pop |>
  filter(year %% 10 == 0) |>
  arrange(scenario, pop_share, mar_growth, year) |>
  group_by(scenario, pop_share, mar_growth) |>
  mutate(
    growth_rate = log(lead(population) / population) / 10,
    year_midpoint = year + 5,
    year_group = paste(year, lead(year) - 1, sep = "-"),
    year_group = fct_reorder(year_group, year_midpoint, max)
  ) |>
  filter(year_midpoint < 300) |>
  ungroup() |>
  select(year_group, year_midpoint, pop_share, mar_growth, scenario, 
         growth_rate)
```


# Check Demographic Measures

```{r}
#| label: fig-life-expectancy
#| fig-cap: Life expectancy at zero for men and women, across all simulations in 25 year periods. A loess smoother is applied to the points. Dashed lines indicate the life expectancy values estimated from the age-specific mortality rates uses as input to the simulation. The first 25 year period does not have anyone at very high ages in order to calculate life expectancy. 

life_expectancy <- life_tables |>
  filter(x == 0) |>
  select(year_group, sex, pop_share, mar_growth, scenario, ex)

life_expectancy |>
  convert_year_group() |>
  ggplot(aes(x = year_midpoint, group = sex, y = ex, color = sex))+
  geom_point(alpha = 0.6)+
  geom_smooth(se = FALSE)+
  geom_hline(yintercept = c(74.1, 79.5), linetype = 2)+
  scale_y_continuous(limits = c(70, 82))+
  scale_x_continuous(breaks = seq(from = 0, to= 300, by = 25))+
  labs(y = "life expectancy at birth", x = "year of simulation")+
  theme_bw()
```

```{r}
#| label: fig-tfr
#| fig-cap: tfr

tfr |>
  convert_year_group() |>
  ggplot(aes(x = year_midpoint, y = tfr))+
  geom_smooth(se = FALSE)+
  geom_point(alpha = 0.6)+
  labs(y = "total fertility rate", x = "year of simulation")+
  scale_y_continuous(limits = c(1.75, 2.5))+
  scale_x_continuous(breaks = seq(from = 0, to= 300, by = 25))+
  theme_bw()
```


```{r}
#| label: fig-growth-rate
#| fig-cap: Population growth rate across simulations at 10 year intervals. GAM smoother shown in blue.

pop_growth |>
  ggplot(aes(x = year_midpoint, y = growth_rate))+
  geom_point(alpha = 0.6)+
  geom_smooth(se = FALSE, method = "gam")+
  scale_y_continuous(labels = scales::percent)+
  scale_x_continuous(breaks = seq(from = 0, to= 300, by = 25))+
  labs(y = "population growth rate", x = "year of simulation")+
  theme_bw()
```

```{r}
sims_pop |>
  filter(scenario == "absorbing state" & mar_growth == "progress" & 
           pop_share == "uneven (85/15)") |>
plot_pop_pyramid(0)

sims_pop |>
  filter(scenario == "absorbing state" & mar_growth == "progress" & 
           pop_share == "uneven (85/15)") |>
plot_pop_pyramid(50)

sims_pop |>
  filter(scenario == "absorbing state" & mar_growth == "progress" & 
           pop_share == "uneven (85/15)") |>
plot_pop_pyramid(75)

sims_pop |>
  filter(scenario == "absorbing state" & mar_growth == "progress" & 
           pop_share == "uneven (85/15)") |>
plot_pop_pyramid(100)
```


- Show observed life expectancy over time for each simulation
- Show observed TFR over time for each simulation
- Show growth rates over time
- Show population pyramids in run in period

## Partnership Modeling

The algorithm for determining who partners with whom is based on reverse-engineering a conditional logit model approach used in the study of assortative mating [@gullickson_counterfactual_2021c]. For each single woman in the population, we sample up to one hundred potential single male partners to construct a potential choice set. We score each potential union of woman $w$ and man $m$ by a log-odds determined as follows:

$$log(O_{mw}) = 0.072 (\textit{age}_m-\textit{age}_w) - 0.014 (\textit{age}_m-\textit{age}_w)^2 + \alpha_{mw}$$

The age difference parameters are derived from @gullickson_counterfactual_2021c and produce a parabolic function of partner age difference that is maximized when the man is 2.57 years older than the woman. The $\alpha$ parameter measures group exogamy where $g_m$ and $g_w$ are the groups for the man and woman, respectively, as follows:

$$\alpha_{mw} = \begin{cases}
  0 & \text{if } g_m = g_w\\
  \alpha & \text{if } (g_m = A \text{ and } g_w = B) \text{ or } (g_m = B \text{ and } g_w = A)\\
  \alpha/2 & \text{else} 
  \end{cases}$$

When both potential partners are from the same group, this parameter is zero. For intergroup partnering between the primary groups of A and B, we use a value of $\alpha$ and for intergroup partnering between A-AB or B-AB, we use half that value. 

We allow $\alpha$ to vary over time within the simulation. Empirically, we determined that the $\alpha$ value roughly corresponds to half the log-odds ratio formed by the cross-product of the two-way table partners' group membership.[^twoway] We use this feature to benchmark a starting $\alpha$ of -5, which corresponds to a log-odds ratio of -10 and an odds of intergroup partnership of less than 0.0001. We then allow $\alpha$ to linearly decline over time, which indicates an exponential increase in the odds of intergroup partnership. For the "progress" scenario, this linear decline continues until $\alpha = 0$ after 100 years, at which point $\alpha$ remains at zero. For the "plateau" scenario, this linear decline continues until $\alpha$ is halved (-2.5) after 50 years and then remains constant. Thus, for the first fifty years of the simulations, the two intergroup partnership scenarios are identical. 

[^twoway]: Formally, if $F_{ij}$ is the frequency of new marriages where $i$ is the husband's group and $j$ is the wife's group, the log-odds ratio is given by $\log(\frac{F_{AB}F_{BA}}{F_{AA}F_{BB}})$.

To determine the actual union chosen, we sample from the available possibilities with weight equal to $O_{mw}$. This approach is equivalent to a conditional logit model approach in which the probability $p_i$ for a specific match $i$ from the available choice set would be given by:

$$p_i = \frac{e^{O_i}}{\sum_{k=1}^{J} e^{O_j}}$$

Using this routine, it is possible for the same male partner to be chosen by two different women. In such cases, we remove duplicate selections randomly. We also found that it was necessary to set a benchmark for situations in which none of the potential male partners crossed a reasonable threshold for acceptability. In cases where women were left with four or fewer options or the highest odds ratio was below 0.12, we terminated the search for a partner and these women remain single for the next year. 

When some groups are very small, sampling may lead to less intergroup partnering than implied by $\alpha$ because the small group will often not be sampled at all within the choice set of a given woman. To avoid this problem, we use stratified sampling of the groups, with a minimum of one member of each group (if greater than zero) in a given choice set.

```{r}
#| label: fig-marriage-lor
#| fig-cap: Estimated log odds ratio of an A-B union over time in five year intervals, across scenario types. Smoothed blue lines are estimated across classification scenarios via a Generalized Additive Model (GAM). Red dashed lines show twice the $\alpha$ parameter for each year.

mar_summary <- sims_mar |>
  filter(year >= years_runin) |>
  mutate(year = year - years_runin) |>
  mutate(year5 = floor(year/5)*5+2.5) |>
  group_by(year5, pop_share, mar_growth, scenario) |>
  summarize(f_aa = sum(hgroup == "A" & wgroup == "A"),
            f_bb = sum(hgroup == "B" & wgroup == "B"),
            f_ab = sum(hgroup == "A" & wgroup == "B"),
            f_ba = sum(hgroup == "B" & wgroup == "A"),
            lor_ab = log(f_ab) + log(f_ba) - log(f_aa) - log(f_bb),
            out_a = (f_ab+f_ba)/(2*f_aa+f_ab+f_ba),
            out_b = (f_ab+f_ba)/(2*f_bb+f_ab+f_ba),
            .groups = "drop") |>
  mutate(
    # remove cases with zeroes somewhere
    lor_ab = if_else(
      is.na(lor_ab) | lor_ab == Inf | lor_ab == -Inf, 
      NA, lor_ab),
    # add alpha value at each year
    alpha = if_else(
      mar_growth == "progress",
      if_else(year5 > 100, 0, -5 - -5 * (year5/100)),
      if_else(year5 > 50, -2.5, -5 - -5 * (year5/100))
    )
  ) |>
  select(year5, pop_share, mar_growth, scenario, starts_with("lor"), 
         starts_with("out"), alpha)

mar_summary |>
  ggplot(aes(x = year5 + 100, y = lor_ab))+
  geom_hline(yintercept = 0, linetype = 3)+
  geom_point(alpha = 0.3, aes(shape = scenario))+
  geom_smooth(se = FALSE, method = "gam")+
  geom_line(aes(y = 2 * alpha), linetype = 2, color = "red", size = 0.75)+
  facet_grid(pop_share~mar_growth)+
  scale_shape_manual(values = c(19, 15, 17, 18, 3))+
  scale_y_continuous(limits = c(-11, 1))+
  labs(y = "log odds ratio of intergroup union",
       x = "year of union",
       shape = "classification scenario")+
  theme_bw()
```

@fig-marriage-lor shows the estimated log-odds ratios of an A-B integroup union over time across all simulations, separated by scenario types. For reference, twice the $\alpha$ value is shown by a red dashed line. The progress and plateau patterns are behaving exactly as expected and we observe no differences in the trend by the starting group distribution. The overall trend within each panel closely matches what was expected based on the $\alpha$ parameter. The only noticeable deviation from expectation is for the uneven starting group size scenario paired with the progress scenario. In this case, the empirical log-odds ratio has a tendency to fall somewhat below the expected value of zero near the end of the simulation. This deviation is a result of the very small size of group B in these scenarios, as discussed in the main paper. When a group is very small, the sampling procedure used to to determine potential partners can break down because the same small set of members are sampled across many possible partnerships. This bias is relatively small except for some years of the absorbing state scenarios. 

Importantly, the $\alpha$ parameters are indifferent to group size differences, which allows us to provide identical parameters across scenarios with different group sizes. The proportion of each group who will partner with members of another group, however, will vary substantially by group size even when $\alpha$ is held constant. @fig-marriage-out shows the proportion of A and B partners who who partner with a member of the other primary group (B and A, respectively) across time and simulation scenario. For the case of even population size, the these proportions are similar for A and B individuals because the group sizes are similar. For the uneven group sizes, however, the proportion is much higher for minority B individuals because they have many more options for partnership with group A than group A members have with group B.

```{r}
#| label: fig-marriage-out
#| fig-cap: Estimated proportion of partners from groups A and B partnered with a member of the other primary group (B and A, respectively) over time in five year intervals, across scenario types. Smoothed blue lines are estimated across classification scenarios via a Generalized Additive Model (GAM).

mar_summary |>
  pivot_longer(cols = starts_with("out"), 
               names_to = "group", 
               values_to = "out") |>
  mutate(group = factor(group, 
                        levels = c("out_a", "out_b"), 
                        labels = c("A", "B"))) |>
  ggplot(aes(x = year5 + 100, y = out, group = group, color = group))+
  geom_point(alpha = 0.3, aes(shape = scenario))+
  geom_smooth(se = FALSE, method = "gam")+
  facet_grid(pop_share~mar_growth)+
  scale_shape_manual(values = c(19, 15, 17, 18, 3))+
  scale_color_manual(values = palette_group)+
  labs(y = "log odds ratio of intergroup union",
       x = "year of union", shape = "classification\nscenario")+
  theme_bw()
```

Notably, the proportion of outgroup partnering declines for the majority group A and increases for minority group B in the uneven group size scenarios in the last 100 years of the simulation, despite the fact that $\alpha$ is held constant during this period. This decline is a result of a decline in the relative size of the group B population making intergroup partnering more difficult for group A, but easier for group B. Some differences across classification scenario types are also driven by underlying group differences, or greater sampling variability when groups become very small (as in the absorbing state case).

```{r}
#| label: fig-marriage-age
#| fig-cap: Kernel density plots of the distribution of partner age at marriage, pooled across all simulations. Densities are shown separately for within-group unions and between-group unions.

sims_mar |>
  pivot_longer(cols = ends_with("age"), 
               values_to = "age", names_to = "spouse") |>
  mutate(type = if_else(hgroup == wgroup, "within group", "between group"),
         spouse = factor(spouse, 
                         levels = c("hage", "wage"), 
                         labels = c("husband", "wife"))) |>
  ggplot(aes(x = age, y = after_stat(density), group = type, fill = type))+
  geom_density(alpha = 0.7, color = NA)+
  facet_wrap(~spouse)+
  scale_fill_manual(values = c("tomato", "navy"))+
  labs(x = "age at marriage", fill = "type of marriage")+
  theme_bw()
```

```{r}
#| label: tbl-age-marriage
#| tbl-cap: Mean age at marriage across simulations for husbands and wives. Results are shown separately for within-group and between-group unions.

sims_mar |>
  mutate(type = if_else(hgroup == wgroup, "w", "b")) |>
  group_by(pop_share, mar_growth, scenario, type) |>
  summarize(mean_age_h = mean(hage),
            mean_age_w = mean(wage),
            .groups = "drop") |>
  pivot_wider(names_from = type, values_from = starts_with("mean_age")) |>
  select(pop_share, mar_growth, scenario, mean_age_h_w, mean_age_h_b, 
         mean_age_w_w, mean_age_w_b) |>
  gt() |>
  cols_label(
    pop_share ~ "Group size",
    mar_growth ~ "Partnering",
    scenario ~ "Classification",
    mean_age_h_w ~ "Within",
    mean_age_h_b ~ "Between",
    mean_age_w_w ~ "Within",
    mean_age_w_b ~ "Between"
  ) |>
  tab_spanner(starts_with("mean_age_h"), label = "Husband") |>
  tab_spanner(starts_with("mean_age_w"), label = "Wife") |>
  fmt_number(starts_with("mean_age"), decimals = 1)
```

`r round(mean(sims_mar$hage-sims_mar$wage), 2)`

## Group Assignment Modeling

A multinomial model used to determine group assignment as follows: 

$$\log(O_{ij}) = \beta_{0j} + \beta_{1j}x_i + \beta_{2j}x_i^2 + \beta_{3j}x_i^3$$ 

where $O_{ij}$ is the odds that individual $i$ is assigned to group $j$ and $x_i$ is the proportion of ancestors of individual $i$ that belong to group A (e.g. a "first generation" child of A and B parents would have 0.5 group A ancestors). We use a third degree polynomial function to allow for non-linearity in this function.

To determine the probability of group assignment from this model, we use:

$$P_{ij} = \frac{e^{O_{ij}}}{\sum_{k=1}^{J} e^{O_{ij}}}$$ 

where $P_{ij}$ is the probability of group assignment to group $j$. We then sample a group assignment for each individual based on these probabilities.

This model requires the estimation of 12 distinct $\beta$ parameters (three slopes and an intercept for each group $j$). To estimate these parameters, we determine probability of group identification at five points along the curve for each group that correspond to different generational loci. We then use these points to empirically estimate the full parameters of the multinomial logit model via maximum likelihood estimation.

<!-- Show a table of our x and y values -->


```{r}
tibble(
  prop_a = c(0.5, 0.75, 0.25, 0.875, 0.125),
  gen_locus = c("1st Generation", rep("2nd Generation", 2), 
                rep("3rd Generation", 2)),
  p_ab_high = c(0.75, 0.675, 0.675, )
)

```


| Proportion of A ancestry | Generational Locus | P(AB) | P(A) | P(B) |
|--------------------------|--------------------|-------|------|------|
| 0.5                      | 1st Generation     | 0.75  | 0.125|0.125 |
| 0.75                     | 2nd Generation     | 0.675 |
| 0.25                     | 2nd Generation     | 0.675 |
| 0.875                    | 3rd Generation     |
| 0.125                    | 3rd Generation     |

The first point we consider is the case of an individual with equal ancestry from groups A and B. This case corresponds to the concept of a "first generation" biracial individual. The probability of AB identification for this case determines the "peak" values we use for the scenario. For individuals who do not identify as AB, we give them an equally likely probability of choosing group A or B.

The next two points we consider are consistent with a "second generation" multiracial individual who has one AB parent and a parent from either A or B. This person would have 75% ancestry from one group and 25% from the other group. For these cases, we determine how much the probability of identifying as AB declines in percentage terms from the peak. In these two cases, we also have to decide how to divide up the probability between single race responses. We assume that the group whose ancestry is more common will be 10 times more likely to be chosen than the group whose ancestry is less common.

The final two points correspond to a "third generation" multiracial individual who has seven great grandparents from one group and one great grandparent from the other group. We again determine how much the probability of AB identification would decline from the second generation case. For single race identification, we assume that the group whose ancestry is more common will be 50 times more likely to be chosen than the group whose ancestry is less common.

Determining the two ratios to use for the second and third generation requires some guess work. The closest empirical work on this topic is @bratter_will_2007b who used American Community Survey data to examine the racial classification of children where one or both parents identified as multiracial. Because we don't know the generational locus of the multiracial parent, we can't determine the generational loci of these children but many will likely be second or third generation. In cases where the multiracial parent was partnered with a single race parent with a racial overlap, the ratios favoring single race identification with the overlapping race ranged from about 60 to 9 [@bratter_will_2007b, p. 833]. Thus, our ratios of 10 and 50, may understate the degree to which single race identification follows the more common ancestry in later generations.

- Show each fit along with points used for fitting.
- Show each fit along with estimated probabilities from actual simulations.

# Ternary Plots

```{r}
#| label: fig-ternary-plot-equilateral
#| fig-cap: Ternary plot showing the distribution of groups A, B, and AB over time. Group distributions are shown by dots at twenty year intervals for each scenario. 
#| fig-height: 6.5
#| fig-width: 6.5

total_pop_group |>
  mutate(year = year - years_runin) |>
  filter(year > 0 & (year %% 10) == 0) |>
  select(-population) |>
  pivot_wider(values_from = share, names_from = group) |>
  arrange(scenario, pop_share, mar_growth, year) |>
  ggtern(aes(x = A, y = AB, z = B))+
  geom_point(aes(color = scenario, group = scenario),
             size = 1.5, alpha = 0.7)+
  geom_path(aes(color = scenario, group = scenario), alpha = 0.7)+
  geom_path(data = tibble(A = c(0.5, 0.25, 0),
                          B = c(0.5, 0.25, 0),
                          AB =c (0, 0.5, 1)),
            linetype = 2)+
  scale_x_continuous(limits = c(0, 1))+
  scale_y_continuous(limits = c(0, 1))+
  facet_grid(pop_share ~ mar_growth)+
  labs(x = "A", y = "AB", z = "B")+
  scale_color_manual(values = palette_scenario)+
  theme_bw()+
  theme(legend.position = "bottom")+
  guides(color = guide_legend(nrow = 2))
```


